---
title: "Lab 01 Week 05 Worksheet"
output: 
  pdf_document:
    keep_tex: true
geometry: margin=1in
fontsize: 11pt
header-includes:
  - \usepackage{setspace}
  - \setlength{\parindent}{0pt}
---

# **R Functions Glossary**

This glossary provides an overview of key R functions used in **Week 5 Lab**, explaining their **purpose** and **general use** in handling time series data.

---

## **1. Downloading Data**

### **`tq_get(symbol, get = "data_source", from = "YYYY-MM-DD", to = "YYYY-MM-DD")`**  
**Purpose:** Retrieves time series data from FRED, Yahoo Finance, or other sources.  
**Example:**  
```{r}
#| eval: false
data <- tq_get("AAPL", get = "stock.prices", from = "2020-01-01", to = "2024-02-28")
```

---

## **2. Handling Dates**

### **`as.numeric(Sys.Date())`**  
**Purpose:** Converts the current date into a numeric value based on the R epoch (January 1, 1970).  
**Example:**  
```{r}
#| eval: false
as.numeric(Sys.Date())
```

### **`ymd("YYYY-MM-DD")`**  
**Purpose:** Converts a character string into a date format.  
**Example:**  
```{r}
#| eval: false
date <- ymd("2024-02-28")
```

---

## **3. Modifying Data**

### **`mutate(dataframe, new_column = operation)`**  
**Purpose:** Adds or transforms columns in a dataframe.  
**Example:**  
```{r}
#| eval: false
data <- mutate(data, price_change = price - lag(price))
```

### **`case_when()`**  
**Purpose:** Recodes values based on conditions.  
**Example:**  
```{r}
#| eval: false
data <- mutate(data, category = case_when(price > 100 ~ "High", TRUE ~ "Low"))
```

---

## **4. Exporting Data**

### **`write_csv(dataframe, "filename.csv")`**  
**Purpose:** Saves a dataframe as a CSV file for use in Tableau.  
**Example:**  
```{r}
#| eval: false
write_csv(data, "time_series_data.csv")
```

---

## **Using This Glossary**
- Reference this list while working through Week 5 Lab.
- Experiment with each function in R to understand how it works.
- Use piping (`%>%`) to combine multiple functions and streamline analysis.
